Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario

### 2021-04-24 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton-title-page.Rmd
Added link to anonymous drive folder to share documents for review purposes; created title page

---
title: "Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario"
author:
  - name: Author One
    email: author1@example.com
    affiliation: Some University
  - name: Author Two
    email: author2@example.com
    affiliation: Another University
    footnote: 1
  - name: Author Three
    email: author3@example.com
    affiliation: Some University
  - name: Author Four
    email: author4@example.com
    affiliation: Some University
address:
  - code: Some University
    address: Department, Street, City, State, Zip
  - code: Another University
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  In this paper we analyze the changes in accessibility to food banks and related services before and during the COVID-19 pandemic in the City of Hamilton, Ontario. Food banks and services are the last line of support for households facing food insecurity; as such, they have become essential and their relevance cannot be ignored in the midst of the economic upheaval caused by the pandemic. Our analysis is based on the application of balanced floating catchment areas, and concentrates on households with lower incomes ($<CAD40,000, approximately the Low Income Cutoff Value for a city of Hamilton's size). Accessibility was low to begin with in suburban and exurban parts of the city. We find that about 14\% of locations originally available in Hamilton closed during the pandemic; the impact of those closures on the level of service and accessibility was disproportionate, with system-wide losses exceeding 40\%. Those losses were geographically and demographically uneven. While every part of the city faced a reduction in accessibility, inner suburbs fared worse in terms of loss of accessibility. As well, children (age $\le 18$) appear to have been impacted the most.

@@

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE, out.width = "100%", align = 'center', fig.width = 7, fig.height = 7
)
```

```{r install-data-package, include = FALSE}
# Run only once if needed to install data package 
if (!require("foodbanksHamilton", character.only = TRUE)) {
  devtools::install_github("https://github.com/paezha/Accessibility-Food-Banks-Hamilton", subdir = "foodbanksHamilton")
}
```

```{r load-packages, cache=FALSE, include=FALSE}
library(foodbanksHamilton) # Data package
library(kableExtra)
library(patchwork)
library(sf)
library(spdep)
library(tidyverse)
#library(tmap)
```

```{r invoke-data, include=FALSE}
data(data_da_2016)
data(dwelling_network_points_2016)
data(foodbank_locations)
data(foodbank_table)
data(hamilton_boundary)
data(modes_less40k)
data(population_age_income)
data(ttm_car)
data(ttm_transit)
data(ttm_walk)
```

```{r remove-irrelevant-destinations, include=FALSE}
## Data preparation
# Remove entries in travel time matrices for inapplicable destinations
# As a first step, we will remove the computed travel times to the three Community Kitchen locations (IDs 1-3):
ttm_car <- ttm_car %>%
  filter(!between(OBJECTID, left = 1, right = 3))

ttm_transit <- ttm_transit %>%
  filter(!between(OBJECTID, left = 1, right = 3))

ttm_walk <- ttm_walk %>%
  filter(!between(OBJECTID, left = 1, right = 3))
```

```{r join-low-income-households, include=FALSE}
## Data preparation
# Join proportion of low income households per DA to dwellings and total number of residential units:
dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
  left_join(data_da_2016 %>%
              st_drop_geometry() %>%
              dplyr::transmute(DAUID = GeoUID, 
                               p_inc_less40k),
            by = "DAUID")
```

```{r probability-of-low-income-per-parcel, include=FALSE}
## Data preparation
# Calculate probability that a point on the network will have low income households. This is the proportion of low income households in the DA multiplied by the total number of residential units at the point:
dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
  mutate(income_less40k = p_inc_less40k * Dwellings)
```

```{r number-of-residential-units-per-DA, include=FALSE}
## Data preparation
# Prepare and join population data from Transportation Tomorrow Survey
# Find number of residential units in each DA. This is needed to assign the population to parcels
residential_units <- dwelling_network_points_2016 %>%
  st_drop_geometry() %>%
  group_by(TAZUID) %>%
  summarize(residential_units = sum(Dwellings))
```

```{r join-population-to-parcels, include=FALSE}
## Data preparation
# Join number of low income households per DA to dwellings and total number of residential units:
dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
  left_join(population_age_income %>%
              st_drop_geometry() %>%
              dplyr::transmute(TAZUID,
                               children,
                               adults,
                               seniors),
            by = "TAZUID") %>%
  left_join(residential_units,
            by = "TAZUID")
```

```{r calculate-low-income-households-per-parcel, include=FALSE}
## Data preparation
# Calculate population per point on the network. This is the population in the DA divided by the total number of residential units. This gives the average number of people per residential unit. Then, multiply by the number of dwellings (residential units) at the point:
dwelling_network_points_2016 <- dwelling_network_points_2016 %>%
  mutate(children = (children/residential_units) * Dwellings,
         adults = (adults/residential_units) * Dwellings,
         seniors = (seniors/residential_units) * Dwellings)
```

```{r join-household-data-to-travel-time-tables, include=FALSE}
## Data preparation
# Join the household data to the travel time matrices:
ttm_car <- ttm_car %>%
  left_join(dwelling_network_points_2016 %>%
              st_drop_geometry() %>%
              dplyr::select(ID, 
                            income_less40k,
                            children,
                            adults,
                            seniors,
                            DAUID, 
                            TAZUID),
            by = c("UID" = "ID"))

ttm_transit <- ttm_transit %>%
  left_join(dwelling_network_points_2016 %>%
              st_drop_geometry() %>%
              dplyr::select(ID, 
                            income_less40k,
                            children,
                            adults,
                            seniors,
                            DAUID, 
                            TAZUID),
            by = c("UID" = "ID"))

ttm_walk <- ttm_walk %>%
  left_join(dwelling_network_points_2016 %>%
              st_drop_geometry() %>%
              dplyr::select(ID, 
                            income_less40k,
                            children,
                            adults,
                            seniors,
                            DAUID, 
                            TAZUID),
            by = c("UID" = "ID"))
```

```{r join-foodbank-data-to-travel-time-tables, include=FALSE}
## Data preparation
# Join the foodbank data to the travel time tables:
ttm_car <- ttm_car %>%
  left_join(foodbank_locations %>%
              st_drop_geometry() %>%
              dplyr::select(ID, 
                            TYPE,
                            COVIDClose),
            by = c("OBJECTID" = "ID"))

ttm_transit <- ttm_transit %>%
  left_join(foodbank_locations %>%
              st_drop_geometry() %>%
              dplyr::select(ID, 
                            TYPE,
                            COVIDClose),
            by = c("OBJECTID" = "ID"))

ttm_walk <- ttm_walk %>%
  left_join(foodbank_locations %>%
              st_drop_geometry() %>%
              dplyr::select(ID,
                            TYPE,
                            COVIDClose),
            by = c("OBJECTID" = "ID"))
```

```{r drop-nas-from-travel-time-tables, include=FALSE}
## Data preparation
# Drop NAs:
ttm_car <- ttm_car %>%
  drop_na(TAZUID,
          income_less40k)

ttm_transit <- ttm_transit %>%
  drop_na(TAZUID,
          income_less40k)

ttm_walk <- ttm_walk %>%
  drop_na(TAZUID,
          income_less40k)
```

```{r proportion-of-trips-by-mode-by-taz, include=FALSE}
## Data preparation
# Calculate proportion of trips by mode per TAZ:
modes <- modes_less40k %>%
  mutate(p_car = Driver/(Driver + Transit + Walk),
         p_transit = Transit/(Driver + Transit + Walk),
         p_walk = Walk/(Driver + Transit + Walk))
```

```{r fill-nas-transport-modes, include=FALSE}
## Data preparation
# Assume that TAZ with NAs have a proportion of car trips of one:
modes <- modes %>%
  dplyr::mutate(p_car = replace_na(p_car, 1),
                p_transit = replace_na(p_transit, 0),
                p_walk = replace_na(p_walk, 0))
```

```{r join-trips-by-mode-to-travel-time-tables, include=FALSE}
## Data preparation
# Join trip-making modal proportions
ttm_car <- ttm_car %>%
  left_join(modes %>%
              st_drop_geometry() %>%
              transmute(TAZUID,
                        p_car,
                        p_transit,
                        p_walk),
            by = "TAZUID")

ttm_transit <- ttm_transit %>%
  left_join(modes %>%
              st_drop_geometry() %>%
              transmute(TAZUID,
                        p_car,
                        p_transit,
                        p_walk),
            by = "TAZUID")

ttm_walk <- ttm_walk %>%
  left_join(modes %>%
              st_drop_geometry() %>%
              transmute(TAZUID,
                        p_car,
                        p_transit,
                        p_walk),
            by = "TAZUID")
```

```{r join-travel-time-tables, include=FALSE}
## Data preparation
# Join Tables into Single TTM
ttm <- ttm_car %>%
  rename(travel_time_car = travel_time) %>%
  full_join(ttm_transit %>%
              transmute(UID,
                        OBJECTID,
                        travel_time_transit = travel_time),
            by = c("UID", "OBJECTID")) %>%
  full_join(ttm_walk %>%
              transmute(UID,
                        OBJECTID,
                        travel_time_walk = travel_time),
            by = c("UID", "OBJECTID"))
```

```{r travel-time-tables-data-check, include=FALSE}
## Data preparation
# Check the NAs in travel time table:
ttm %>% 
  filter(is.na(travel_time_transit)) %>%
  select(travel_time_transit,
         p_transit) %>%
  summary()

# What do NAs here mean? Most DAs missing travel time by transit do not report trips by transit in the TTS. In some DAs there are trips by transit (p_transit != 0), but no foodbank can be reached by transit (travel_time_transit is NA). How many of these cases?
ttm %>% 
  filter(is.na(travel_time_transit) & p_transit > 0) %>%
  select(travel_time_transit,
         p_transit) %>%
  summary()

# It should be fine to replace the NAs with zeros, which multiplied by p_transit will return a zero (i.e., it is not possible to reach a foodbank by transit).
ttm <- ttm %>% 
  mutate(travel_time_transit = replace_na(travel_time_transit, 0))

# Check the NAs in travel time by walk:
ttm %>% 
  filter(is.na(travel_time_walk)) %>%
  select(travel_time_walk,
         p_walk) %>%
  summary()

# Most DAs missing travel time by walking do not report trips by walking in the TTS. In some DAs there are trips by walking (p_walk != 0), but no foodbank can be reached by walk (travel_time_walk is NA). How many of these cases?
ttm %>% 
  filter(is.na(travel_time_walk) & p_walk > 0) %>%
  select(travel_time_walk,
         p_walk) %>%
  summary()

# Like with transit, it should be fine to replace the NAs with zeros, which multiplied by p_walk will return a zero (i.e., it is not possible to reach a foodbank by walking).
ttm <- ttm %>% 
  mutate(travel_time_walk = replace_na(travel_time_walk, 0))

#Summary of the table after these operations:
summary(ttm)
```

```{r calculate-expected-travel-times, include=FALSE}
## Data preparation
# Calculate expected travel times by mode. These are the travel time by the mode multiplied by the probability of using the mode (which is proxied by the proportion of trips by mode in the DA from the TTS). The total expected travel time is the sum of the expected travel times by mode:
ttm <- ttm %>%
  mutate(e_travel_time_car = travel_time_car * p_car,
         e_travel_time_transit = travel_time_transit * p_transit,
         e_travel_time_walk = travel_time_walk * p_walk,
         e_travel_time = e_travel_time_car + e_travel_time_transit + e_travel_time_walk)
```

\newpage

Introduction
==========================

Food insecurity is defined as an "inadequate or uncertain access to a sufficient quantity and/or adequate quality of food" due to a household's financial limitations [@enns2020experiences]. This condition has been associated with reductions in nutritional outcomes [@bhattacharya2004poverty; @kirkpatrick2008food; @olson1999nutrition] and physical and mental health in children and adults [@elgar2021relative; @jones2017food; @ramsey2011food; @seligman2010food; @stuff2004household]. Over at least the past four decades food banks and related services have become an essential line of defense against food insecurity in Canadian communities [@riches2002food; @holmes2018nothing; @black2020examining; @tarasuk2020relationship]. In this respect, Canada is not unlike numerous other wealthy countries where a systematic dismantling of the welfare state took place in the intervening period [@tarasuk2014food].

The emergence of COVID-19, the worst public health crisis since the 1918 flu pandemic, has revealed important social and economic fault lines, and pre-existing patterns of inequality appear to have been exacerbated. Along several other dimensions of stress [e.g., accessibility to health care facilities, @pereira2021geographic], this seems to be the case for food insecurity as well [@laborde2020poverty]. In the US, for example, it has been estimated that there was an increase of more than 30\% in household food insecurity, and more than one third of households were discovered to be newly food insecure - meaning they did not experience food insecurity before the pandemic [@niles2020early]. In Canada, Men and Tarasuk [-@men2021food] report that about 25\% of individuals who experienced job insecurity (a relatively common occurrence during the pandemic), also experienced food insecurity. Similarly, according to Statistics Canada [-@statisticscanada2020food], in the early stages of the pandemic almost 15\% of individuals reported living in a household that faced food insecurity; the risk of food insecurity was substantially higher for households with children. The difference between households with and without children was significant, and 11.7\% of households with children indicated that "food didn’t last and [there was] no money to get more" sometimes or often, compared to 7.3\% of households without children; likewise, 13\% of households with children indicated that they "[c]ouldn’t afford balanced meals" sometimes or often, compared to 8.8\% of households without children.

The impacts of food insecurity during the pandemic are alarming, since diet-related diseases, such as obesity, heart-disease, and diabetes, were already critical public health concerns in Canada prior to COVID-19 [@boucher2017ontario]. While food banks are not necessarily a stable solution to food insecurity and in fact may encourage a retrenchment of neoliberal policy [@wakefield2013sweet], at least they can be argued to provide a resource of last instance to households in precarious situations [@bazerghi2016role]. As recently as 2019, the Hamilton Hunger Report^[https://www.hamiltonfoodshare.org/wp-content/uploads/Hamilton-Food-Share-Hunger-Report-2019.pdf] noted that food banks in Hamilton, Ontario, recorded the highest number of visitors in the past 29 years; the number of children visiting food banks (minors up to 18 years old) was 9,125 in March 2019, up from 8,278 the year before, a rate of increase greater than population growth.

It is known that urban food environments, within which people make their daily food choices, are essential in influencing eating behaviours and health outcomes, based on factors such as food availability, ease of geographic accessibility and socio-demographic variations [@paez2010relative; @vanderlee2017food; @widener2018spatial]. However, while there is a wealth of literature that has examined the topic of geographic accessibility to healthy food through the "food desert" concept, there has been little research into accessibility to food banks. Although previous work has explored differences in _accessing_ food banks, such as how some households utilize food banks over short periods of time while others regularly utilize food banks as longer-term resource [e.g., @enns2020experiences], we are not aware of any research that has focused on the geographic component of accessibility.

The study of place-based geographic accessibility is concerned with capturing the potential to reach destinations of value using the transportation network [@paez2012measuring]. Indeed, the Government of Canada's recent Food Policy has made "access" to healthy food a priority for Canadian communities^[https://www.agr.gc.ca/eng/about-our-department/key-departmental-initiatives/food-policy/the-food-policy-for-canada/?id=1597863791042] and previous survey research has suggested that such accessibility plays a key role in user satisfaction with food bank service delivery [@holmes2018nothing]. However, as with research into the prevalence of food deserts, accessibility to food banks is unlikely to be evenly distributed, and variation throughout a city can be expected due to transportation network characteristics, and the spatial distribution of food bank locations and the population they are meant to serve. Furthermore, policy responses to the COVID-19 pandemic likely have added to the distress of vulnerable households. Non-pharmaceutical interventions during the pandemic involving restrictions in mobility have increased the friction of travel, in particular by transit on which low income populations are more reliant [e.g., @deweese2020tale]. At the same time, the pandemic has created additional stress for the operators of food banks through disruptions in the supply chain [e.g., @mckay2021exploring] as well as concerns surrounding the delivery of service in safe conditions and possible cancellation of food service programs.

For this study, we aim to look at how the landscape of food banks and related services (e.g. low-cost or free meal service providers) available in Hamilton, Ontario, changed during the pandemic. Did the number of open food bank services diminish? If so, what was the accessibility to food banks before the pandemic from the perspective of low income households, and how it changed during the pandemic? And finally, who are most likely to have been impacted by changes in the accessibility landscape? This paper first looks at the distribution of food banks and related services before and during the pandemic. Then, we use the balanced floating catchment area approach of Paez et al. [-@paez2019demand] to investigate the accessibility situation. For this, we use a fully disaggregated approach based on parcel-level data. Socio-economic and demographic data are drawn from the latest Census of Canada (2016), whereas travel information is from the most recent regional travel survey from 2016. This paper follows reproducible research recommendations [see @brunsdon2020opening], and the research was conducted using open source tools for transportation analysis [@lovelace2021open]. The code and data necessary to reproduce the analysis are available in a public repository^[add repository].

Food Insecurity and Food Bank Use in Canada
============

Food insecurity is the inability to acquire and consume an adequate amount or good quality food, leading to inadequate nutrient intake [@enns2020experiences; @kirkpatrick2008food; @tarasuk2009household]. This nutrient deficiency is associated with major population health concerns, particularly among Canadians at socio-economic disadvantage [@bazerghi2016role]. Data on food insecurity in Canada is collected in several ways. Quantitatively, official government surveys such as the Household Food Security Survey Module (HFSSM), the Canadian Community Health Surveys (CCHS), the International Study of Adults (LISA), and official classifications determined by Health Canada in relation to socio-demographic variables offer some insight into food insecurity [@kirkpatrick2008food; @tarasuk2009household; @gundersen2018food]. Nationally, it has been found that food insecurity impacts approximately 12.3% of Canadian households [@tarasuk2014food]

Food banks - sometimes also referred to as ‘food pantries’ and ‘food shelves’ - originated as a community response to aid those with inadequate food by voluntarily offering them meals and ingredients [@riches2002food; @loopstra2012relationship]. The scope and objectives of food banks can vary by region and by country, and these organizations can include not only prepared meals and aliments, but also shared spaces to connect in community gardens and community kitchens [@wakefield2013sweet]. Although in their origin food banks were meant to be provide a temporary solution to accommodate those in hunger due to job retrenchments and economic downfalls since the 1980s, over time they have evolved into a community practice to secure emergency food supplies for those in need [@loopstra2012relationship; @wakefield2013sweet].

In Canada, the number of food banks has steadily increased in the past few decades [@wakefield2013sweet]. The largest database of food banks and their use comes from the non-profit association Food Banks Canada (FBC), which conducts an annual assessment through its affiliated members. FBC's 2018 Hunger Count report^[https://foodbankscanada.ca/getmedia/241fb659-05f5-44a2-9cef-56f5f51db523/HungerCount-2018_FINAL_EN.pdf.aspx?ext=.pdf] (the most recent available) listed 1,830 member food banks across the country, and found that Canadians visited food banks 1.1 million times in March of 2018. Of those accessing food banks, certain population characteristics tend to be over-represented compared to national totals from the 2016 Canadian Census of Population. According to FBC's 2018 data, single-adult households represent 45% of those utilizing food banks despite making up 28% of Canada's population, 19% are single-parent households (compared to 10% nationally), and 35% of those accessing food bank services are children aged 0-18 even though their share of Canada's national population is approximately 20%. In addition, 59% of households accessing food banks list social or disability assistance as their primary source of income. Similarly, using data from the 2011-2012 CCHS, Tarasuk et al. [-@tarasuk2019geographic] found higher odds of food insecurity amongst households relying on social assistance, those without a university degree or with children under the age of 18, and individuals that lived alone, renters, and those identifying as Aboriginal. While surveys revealed that only 20 to 30 percent of those experiencing food insecurity were found to frequent food banks in Canada [@tarasuk2014food], research from Ottawa [@enns2020experiences] and Vancouver [@black2020examining] suggests that long-term users tend to be older, have health or mobility challenges, live in large households, and are less likely to have employment income.

Economic stress caused by the COVID-19 pandemic, rising unemployment rates, and changes in poverty levels have disrupted the food environment and led to higher rates of food insecurity [@niles2020early]. Pandemic-related food security studies in the US have found a substantial increase in households experiencing food insecurity for the first time, and also in households experiencing more severe food insecurity than before [@niles2020early; @wolfson2020food]. Most recently in May 2020, Canada recorded 14.7% of its population living in food insecurity in the past 30 days [@statisticscanada2020food]. Considering the negative mental and physical health effects associated with food insecurity, increases in food insecurity rates due to the pandemic, signal a change in the food environment with potential damages to population health during the course of the pandemic and beyond [@niles2020early]. Recent data^[https://www.foodbankscanada.ca/FoodBanks/MediaLibrary/COVID-Report_2020/A-Snapshot-of-Food-Banks-in-Canada-and-the-COVID-19-Crisis_Exec-Sum_EN.pdf] from FBC showed that 52% of member food banks reported an increase in usage in March of 2020 when initial lockdown restrictions were put in place across much of the country. The pandemic also created significant staffing issues with 42% of food banks reporting a reduction in volunteers. However, 53% of food banks later reported a decrease in use into the summer of 2020 which FBC members attributed to emergency financial support programs from the federal government. Nevertheless, some of these benefit programs were temporary, which suggests that many households may again turn to food bank services to meet their needs.

In terms of geography, previous research conducted at the provincial scale using data from the 2011-2012 CCHS found that the prevalence of food insecurity ranged across the country from 11.8% of households in Ontario to 41% of households in Nunavut [@tarasuk2019geographic]. However, although previous research has examined the characteristics of individuals and households accessing food banks, the locational or transportation accessibility aspect of food bank access is not well understood. A wealth of literature examining the food desert concept suggests that, in addition to socio-economic and demographic factors, location and transportation networks play a key role in a household's accessibility to healthy foods [@paez2010relative; @vanderlee2017food; @widener2018spatial]. For food banks specifically, previous qualitative research in Ontario by Smith-Carrier et al. [-@smith2017food] has noted that "transportation can be challenging, particularly if the food bank is situated in a remote location" (p. 32). Particularly, it appears that participants experience challenges with the "inordinate amount of time necessary to obtain food, and difficulties associated with transportation" (p. 39). Users of food banks, according to this research, rely on a variety of modes of transportation to access services. Consequently, the location of facilities matters; in the words of an interviewee: "I wish it [the food bank] was a little more centrally located. Because if I didn’t have a bike I’d have to walk it all the way out there and back. I wonder about people who don’t" (p. 39).

To offer greater insight into the role of transportation and location in food bank accessibility, this research examines how geographic accessibility to food banks and food services changed in Hamilton during the COVID-19 pandemic.

Methods and Materials
===================

## Methods

For the research in this paper we adopt the balanced floating catchment area approach of Paez et al. [-@paez2019demand]. This method for estimating accessibility is a form of the widely-used two-stage floating catchment area method [@radke2000spatial; @luo2003measures]. Floating catchment areas are used to estimate accessibility when there are potential congestion effects, and operate by calculating first the _demand_ for spatially distributed services. The demand (usually the number of people who require a service) is used to calculate a level of service. In a second step, the level of service is allocated back to the population. Demand and level of service are allocated using some form of distance-decay to embody the geographical principle that, given a choice, people prefer to travel less than more when reaching destinations.

More formally, the first step of this method is as follows:
$$
L_j = \frac{S_j}{\sum_{i=1}^nP_iw_{ij}}
$$

\noindent where $S_j$ is the level of supply at location $j$, in simplest terms whether a service point is present (i.e., $S_j=1$) or not (i.e., $S_j=0$); $P_i$ is the population at location $i$ that demands the service; and $w_{ij}$ is a weight, typically a function of the distance between locations $i$ and $j$. $L_j$ is the level of service at location $j$ and it is the inverse of the number of people that need to be serviced.

The second step in this process is then summing the level of service that each population unit can reach, according to the distance-decay weight:
$$
A_i = \sum_{j=1}^JL_jw_{ji}
$$

\noindent where $A_i$ is the accessibility to the service, which is in the same units as the level of service: as the inverse of the population being serviced. When the population being serviced is low accessibility is high (i.e., there is little competition for the service), and viceversa.

Floating catchment area methods are prone to overestimation of the population and the level of service due to multiple-counting. The population at $P_i$ is allocated to _every_ service point $j$ for which $w_{ij}>0$. Similarly, the level of service at $LOS_j$ is allocated to _every_ population point for which $w_{ji}>0$. This inflation effect has been known for several years, and several modifications have been proposed to mitigate it [e.g., @wan2012three; @delamater2013spatial]. A definitive solution to this issue was presented by Paez et al. [-@paez2019demand]. In order to avoid the multiple-counting in the summations, the population and the level of service need to be allocated _proportionally_. This is achieved by standardizing the weights as follows:
$$
w_{ij}^\text{st} = \frac{w_{ij}}{\sum_{i=1}^nw_{ij}}
$$
\noindent and:
$$
w_{ji}^\text{st} = \frac{w_{ji}}{\sum_{j=1}^Jw_{ji}}
$$

The standardized weights satisfy the following conditions:
$$
\sum_{i=1}^nw_{ij}^\text{st}=1
$$

\noindent and:
$$
\sum_{j=1}^Jw_{ji}^\text{st}=1
$$

Since the population is allocated proportionally, its value is preserved:
$$
\sum_{i=1}^nP_iw_{ij}^\text{st}=P_i
$$

\noindent as is the level of service:
$$
\sum_{j=1}^JL_jw_{ji}^\text{st}=L_j
$$
## Study Area

**Need to provide some context here: the geography which makes transportation challenging, i.e., due to the escarpment; the main parts of the city: city core, inner suburbs, industrial north, suburban, exurban; and the industrial background of the city maybe, with poverty rates compared to other Canadian cities?**

## Data

Data have been prepared for sharing as a data package^[add repository]. The contents of the data package are described next.

### Statistics Canada

Population and income statistics for 2016 were retrieved at the level of Dissemination Areas (DAs) using the package `cancensus` [@vonBergmann2021cancensus]. DAs are the smallest publicly available census geography in Canada. Income data corresponds to the count of households by different total income groupings.

### Origins: Residential parcels

We converted all recorded residential land parcels in the City of Hamilton to points on the road network. Each point includes information about the number of residential units in the parcel. Next, we define low-income households as those having a total income of less than CAD40,000, which is approximately the mid-point of the low income cut-off (LICOs) for families in Canadian cities with populations greater than 500,000 in 2016, to match other Census data [@statisticscanada2020licos]. We then "populate" each residential unit with the probability of being a low-income household based on the counts of households by income groups in the DA in which the parcel is located. While this method assumes a constant probability of low-income household status for all residential units in a DA, the parcel-level analysis affords a high level of spatial disaggregation for the accessibility analysis.

### Destinations: Food Banks and Food Service Locations

The locations of food banks and related food services were obtained from the Hamilton Public Library's Food Access Guide^[http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf]. The guide was updated in April of 2021 to indicate any change affected on the services due to the pandemic. This includes modified business hours, a need to make reservations before frequenting, and locations that have completely shut down in consequence. Table 1 defines each service type and the number of locations pre- and during the COVID-19 pandemic. While some food bank services have a specific a target population, such as prioritizing families with young children aged between 0 and 3 or accepting only those providing proof of low-income status through housing and utility statements, all the food bank services indicated below are designed to accommodate those in need of food at zero to low cost. With our focus on food banks and food services that offer free or low-cost meals at particular locations, we first removed services such as Meals on Wheels and other food access services such as food box, community kitchens, student nutrition programs, and shopping and transportation. In addition, two free meal services held on different days at the same location were collapsed into a single service point for the accessibility analysis. Additional details on the operations of individual facilities is not publicly available, and with the changes in operations it proved unfeasible to collect it. For this reason, the analysis to follow is of accessibility to the location of food banks and services, but not to specific services (e.g., breakfasts vs. food boxes).

```{r table-food-bank-info, echo=FALSE, results='asis'}
foodbank_table %>% kable("latex",
                         booktabs = TRUE,
                         align = ("cccccc"),
                         col.names = c("Type", 
                                       "Description", 
                                       "Locations Pre-COVID", 
                                       "Locations During COVID", 
                                       "Additional Notes"),
                         caption = "\\label{tab:food-bank-info}foodbank and Food Service Information.",
                         escape = FALSE) %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  column_spec(2, width = "15em") %>%
  column_spec(5, width = "15em")
```

### Routing and travel time tables

Travel time tables for three modes (car, transit, walking) were computed using the parcels as the origins and the locations of the food banks as the destinations. For routing, the package `r5r` [@pereira2021r5r] was used with a network extract for the City of Hamilton from OpenStreetMaps and the General Transit Feed Specification from Hamilton Street Railway, the local transit operator. For routing purposes we used maximum values of 180 min and 10,000 m walking distance: any destination that exceeded these thresholds was ignored. The departure time used for routing was 8:00AM on March 30, 2021 to reflect transit service around the morning service peak on a typical Tuesday.

### Transportation Tomorrow Survey

We used the Data Retrieval System of the Transportation Tomorrow Survey (TTS)\footnote{\url{http://dmg.utoronto.ca/}} to download cross-tabulations of: 1) primary mode of travel per trip by income by place of residence; and 2) age by income by place of residence. These data are from the 2016 Survey (the most recent available). The data are geocoded at the level of Traffic Analysis Zones (TAZ) using the most recent zoning system from 2006 and expansion factors are applied to weight the trips . Each parcel point is populated with the proportion of trips by three modes of travel: car (as driver or passenger), transit, and walk.

### Expected Travel Times

Once we obtained travel time tables with population (number of households) and proportion of trips by mode, we calculated the expected travel time $ett$ from each parcel $i$ to a food bank or food service location $j$ as follows:
$$
ett_{ij} = p^c_i\cdot tt^c_{ij} + p^t_i\cdot tt^t_{ij} + p^w_i\cdot tt^w_{ij}
$$

\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the travel time from parcel $i$ to the food bank. In other words, the expected travel time is the weighted sum of travel times to the food bank, with the weights given by the expected modal split of trips made by low-income households in the TAZ per the TTS data.

Results and Discussion
===================

```{r accessibility-threshold-sensitivity, include=FALSE}
## Analysis
# Testing different thresholds to check how accessibility changes. Here, the thresholds are varied from 2 min to 30 min in 2 min step increments
threshold_seq <- seq(from = 2, to = 30, by = 1)

# Initialize a data frame to store results
accessibility_test <- data.frame(threshold = threshold_seq, 
                                 accessibility = numeric(length(threshold_seq)))

# Calculate the system-wide accessibility according to each threshold
for(i in 1:length(threshold_seq)){
  results_test <-  ttm %>% 
    transmute(UID, OBJECTID, travel_time = e_travel_time, population = income_less40k, supply = 1) %>%
    b2sfca(threshold = i )
  accessibility_test$accessibility[i] <- sum(results_test$accessibility$accessibility)
}
```

```{r results-pre-covid, include=FALSE}
## Analysis 
# Calculate k-min accessibility pre-covid:
results_pre <- ttm %>%
  #dplyr::filter(COVIDClose == "No") %>%
  transmute(UID, 
            OBJECTID, 
            travel_time = e_travel_time, 
            population = income_less40k, 
            supply = 1) %>%
  b2sfca(threshold = 15)
```

```{r results-during-covid, include=FALSE}
## Analysis 
# Calculate k-min accessibility during covid:
results_covid <- ttm %>%
  dplyr::filter(COVIDClose == "No") %>%
  transmute(UID, 
            OBJECTID, 
            travel_time = e_travel_time, 
            population = income_less40k, 
            supply = 1) %>%
  b2sfca(threshold = 15)
```

```{r, bind-accessibility-results, include=FALSE}
## Analysis
# Bind level of service results:
los <- rbind(data.frame(results_pre$los, 
                        timing = "Pre-covid", 
                        threshold = "30 min"),
             data.frame(results_covid$los, 
                        timing = "During covid", 
                        threshold = "30 min")) %>%
  mutate(timing = factor(timing, 
                         levels = c("Pre-covid", 
                                    "During covid"),
                         ordered = TRUE))

# Bind accessibility results:
accessibility <- rbind(data.frame(results_pre$accessibility, 
                                  timing = "Pre-covid", 
                                  threshold = "30 min"),
                       data.frame(results_covid$accessibility, 
                                  timing = "During covid", 
                                  threshold = "30 min")) %>%
  mutate(timing = factor(timing, 
                         levels = c("Pre-covid", 
                                    "During covid"),
                         ordered = TRUE))
```

Figure \ref{fig:foodbanks} shows the location of food banks and services in the City of Hamilton and their status. Before the pandemic there were `r nrow(foodbank_locations)` of which `r nrow(foodbank_locations %>% filter(COVIDClose == "Yes"))` (`r round(nrow(foodbank_locations %>% filter(COVIDClose == "Yes"))/nrow(foodbank_locations) * 100, 2)`\%) closed during the pandemic. As shown in the figure, food services tend to be predominantly located in the central parts of the city. This is not surprising: population density is high there, and it is also the part of the city where lower income households are more numerous in absolute and relative terms (see Figure \ref{fig:low-income-households}). Alas, this is also the part of the city where most of the closures during the pandemic happened.

```{r plot-location-foodbanks, echo=FALSE, fig.cap="\\label{fig:foodbanks}Location of food banks/services and operation status; the dotted box is an inset of the central part of the City of Hamilton"}
## Analysis
# Plot location of foodbanks

# Create a bounding box for an inset
inset_box <- st_bbox(c(xmin = -79.975,
                       xmax = -79.775, 
                       ymax = 43.2, 
                       ymin = 43.3), 
                     crs = st_crs(4326)) %>%
  st_as_sfc()

>>>>

---
title: "Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario"
author:
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: University of Toronto Scarborough
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Gyoorie Kim
    email: gyoorie.kim@mail.utoronto.ca
    affiliation: University of Toronto Mississauga
  - name: Jue Wang
    email: gis.wang@utoronto.ca
    affiliation: University of Toronto Mississauga
address:
  - code: University of Toronto Scarborough
    address: Department of Geography & Planning, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON M1C1A4
  - code: McMaster University
    address: School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada
  - code: University of Toronto Mississauga
    address: Department of Geography, Geomatics and Environment, University of Toronto Mississauga, 3359 Mississauga Road, Mississauga, ON, L5L 1C6 , Canada   
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  In this paper we analyze the changes in accessibility to food banks and related services before and during the COVID-19 pandemic in the City of Hamilton, Ontario. Food banks and services are the last line of support for households facing food insecurity; as such, they have become essential and their relevance cannot be ignored in the midst of the economic upheaval caused by the pandemic. Our analysis is based on the application of balanced floating catchment areas, and concentrates on households with lower incomes ($<CAD40,000, approximately the Low Income Cutoff Value for a city of Hamilton's size). Accessibility was low to begin with in suburban and exurban parts of the city. We find that about 14\% of locations originally available in Hamilton closed during the pandemic; the impact of those closures on the level of service and accessibility was disproportionate, with system-wide losses exceeding 40\%. Those losses were geographically and demographically uneven. While every part of the city faced a reduction in accessibility, inner suburbs fared worse in terms of loss of accessibility. As well, children (age $\le 18$) appear to have been impacted the most.

@@
